---
title: "Roleplaying Game Dice: Statistics and Gameplay"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE}
library(ggplot2)
library(knitr)
```

```{r fate dice, echo=FALSE}
n <- 4
l <- rep(list(-1:1), n)
fate=expand.grid(l)
fate$result=rowSums(fate)
fate$dice="fate"
```

```{r pbta dice, echo=FALSE}
n <- 2
l <- rep(list(1:6), n)
aw=expand.grid(l)
aw$result=rowSums(aw)
aw$dice="pbta"
```

```{r pbta adv, echo=FALSE}

n <- 3
l <- rep(list(1:6), n)
aw_mod=expand.grid(l)
aw_mod$sum=rowSums(aw_mod)
aw_mod$min = do.call(pmin, aw_mod[,1:3])
aw_mod$max = do.call(pmax, aw_mod[,1:3])
aw_mod$adv = aw_mod$sum-aw_mod$min
aw_mod$dis = aw_mod$sum-aw_mod$max

aw_adv = aw_mod[c("adv")]
aw_adv$dice= "pbta_adv"


aw_dis = aw_mod[c("dis")]
aw_dis$dice= "pbta_dis"


```


## Introduction
In 2015, I won a contest from [Evil Hat Productions](https://www.evilhat.com/home/) as part of the Evil Hat Street team. My prize was an hour long conversation with members of the Evil Hat team on a subject of my choosing. My friend [John](https://github.com/owendl/rpgaday2020/blob/master/29-ride-DO.md) and I got on a call with Sean Nittner and Fred Hicks to discuss game design. In our conversation, Sean brought up the balance of agency versus experience in roleplaying games. The interaction of these two concepts with game dice mechanics in two different roleplaying game systems is the subject of this paper.

### Agency
Agency decscribes the control the players have over the game and the characters destiny (Reference for more explanation: [link](http://game-wisdom.com/critical/player-agency-game-design-narrative). Games with high agency give players a greater degree of control of the world and what happens. In these games, the players are exerting their will on the game, primarily via their characters. Other narrative options are available in some games (Assets in Cortex, narrative declaration in Fate) but as these do not participate with the dice mechanics they are not considered.

### Experience
By contrast experience is the world of the game imposing its will on the characters. Low agency games In the context of the original conversation, experience was framed as the inverse of agency. Thus a low agency game, could be considered as high experience and vice versa. 

A high Experience game then sees players with less control over the game world through the dice.

## Drama and Dice

### Powered by the Apocalypse
Powered by the Apocalypse games (PbtA) utilize a dice mechanic first popularized by the game Apocalypse World by Vincent Baker. The governing dice mechanic could not be any simpler, roll two six-sided dice (2d6) and add a stat, typically ranging from -1 to 2. A 6 or lower (6-) is a failure, 7-9 is a partial success and a 10+ is a full success. What is a success and variations thereof can change game to game but the core mechanic stays the same. 

Below is a table of the potential results from this mechancic and their proportion of possible roles (ignoring stat modification).

```{r, echo=FALSE}
kable(t(table(aw$result)/nrow(aw)),digits = 3, caption  = "Frequency of Results of 2d6")
```

A few notes on these results:

* 42% of PbtA rolls are a failure on the dice (6-)
* 42% of PbtA rolls are only a partial success on the dice (7-9)

### Fate
The Fate roleplaying system builds off of the FUDGE roleplaying system dice mechanic, where four FUDGE/Fate dice (4df) are rolled and then the combined results are totaled. These dice are also six-sided but two faces are blank (0), two have a "-" sign (-1) and two faces have a "+" sign (+1). Thus the range of values for a 4df roll is -4 to 4. Beyond the dice themselves, players add skills to the rolls, typically ranging from 0 to 5. 

```{r, echo=FALSE}
kable(t(table(fate$result)/nrow(fate)),digits = 3, caption  = "Frequency of Results of 4df")
```

Consideration of failure or success based on the dice is more difficult to determine because there is not standard target number. However, in the context of control, it is easy to predict the outcome of a roll based on the skill used. Almost two out of three rolls, 63%, of Fate dice rolls are -1, 0 or 1. Thus, most rolls are approximately the value of the character's skill.

### Savage Worlds

### Comparing Different Dice

```{r, message=FALSE, echo=FALSE}
aw$result=aw$result-7
dice=rbind(fate[,c("result", "dice")], aw[,c("result", "dice")])
```

```{r, echo=FALSE}
  ggplot(dice,aes(x=result)) +
  geom_histogram(data=subset(dice,dice == 'fate'),aes(fill = dice,y=..density..), binwidth=1, alpha = 0.2) +
    geom_histogram(data=subset(dice,dice == 'pbta'),aes(fill=dice,y=..density..),binwidth=1,alpha = 0.2)+
  geom_line(stat="function",fun = dnorm, args = list(mean = 0, sd = 2), aes(color="N(0,4)"))+ geom_line(stat="function",fun = dnorm, args = list(mean = 0, sd = 3), aes(color="N(0,9)"))
```

The two additional lines are of two normal distributions $N(0,4), N(0,9)$. These provide rough estimations of the distributions of the two different dice mechanics.

### Agency and Experience: a personal thoughts
In Fate games, the outcomes are more predictable, the dice clustering around zero. Extreme results are possible but very unlikely. Powered by the Apocalypse games have a flatter spread, extreme results entering the realm of plausible. And failure is guaranteed almost half of the time. 

I remember thinking during our conversatoin with Fred and Sean that I play Fate for the agency and Powered by the Apocalypse games for the experience. I didn't think much about it until recently when I began to ask myself what the shape of the dice looks like. It makes sense that the game was called Apocalypse World, the dice are very much the world acting on you. Contrast that with Fate, where the outcome is actually within your grasp (and even more so when considering Fate points).

I find Fate games really sing with a lighter, more playful tone while the best PbtA games lean into the seriousness their names imply. Neither game is really better than the other, each provides its own tenor and style to games.

### Update Jan 2023
I have recently come across a variation on the Apocalypse World dice mechanic called advantage and disadvantage. A player with advantage (something in the narrative favoring them) a player gets to roll three six-sided dice (3d6) and sum the the two highest dice (a roll of 2,4,6 becomes a result of 10). If a player rolls with disadvantage (something in the fiction is acting against them) they again roll 3d6 but sum the two lowest dice (a roll of 2,4,6 becomes a result of 6). I was curious to look at how this affected the dice distributions. Below is the distribution of normal Apocalypse World (pbta), with advantage (pbta_adv) and disadvantage (pbta_dis).

```{r, message=FALSE, echo=FALSE}
aw_adv$result=aw_adv$adv-7
aw_dis$result=aw_dis$dis-7
dice2=rbind(fate[,c("result", "dice")], aw[,c("result", "dice")],aw_adv[,c("result", "dice")],aw_dis[,c("result", "dice")])
```

```{r, echo=FALSE}
  ggplot(dice2,aes(x=result)) +
  # geom_histogram(data=subset(dice,dice == 'fate'),aes(fill = dice,y=..density..), binwidth=1, alpha = 0.2) +
    geom_histogram(data=subset(dice,dice == 'pbta'),aes(fill=dice,y=..density..),binwidth=1,alpha = 0.2)+
  #
  geom_line(stat="function",fun = dnorm, args = list(mean = 0, sd = 3), aes(color="pbta N(0,9)"))+
  geom_line(stat="function",fun = dnorm, args = list(mean = -1.46, sd = 2.22), aes(color="pbta_dis N(-1.46,4.93)"))+ 
  geom_line(stat="function",fun = dnorm, args = list(mean = 1.46, sd = 2.22), aes(color="pbta_adv N(1.46,4.93)"))+
  geom_histogram(data=subset(dice2,dice == 'pbta_adv'),aes(fill=dice,y=..density..),binwidth=1,alpha = 0.2)+
  geom_histogram(data=subset(dice2,dice == 'pbta_dis'),aes(fill=dice,y=..density..),binwidth=1,alpha = 0.2)
```

Intuitively it makes sense that advantage or disadvantage modified the mean of the distributions. What I found more interesting was that the variance also changed drastically, moving into close to the variance of a high agency game like Fate.