---
title: "Roleplaying Game Dice Distributions"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,includee=FALSE}
library(ggplot2)
```


## GitHub Documents
About five years ago, I won a contest from [Evil Hat Productions](https://www.evilhat.com/home/) as part of the Evil Hat Street team. My prize was an hour long conversation with members of the Evil Hat team on a subject of my choosing. To be honest, I was not terribly interested in the conversation at the time (I wanted the tangible goodies from the other prizes). But my friend [John](https://github.com/owendl/rpgaday2020/blob/master/29-ride-DO.md) and I got on a call with Sean Nittner and Fred Hicks to discuss game design. 

## Including Code

You can include R code in the document as follows:

```{r fate dice}
n <- 4
l <- rep(list(-1:1), n)

fate=expand.grid(l)

fate$result=rowSums(fate)
# hist(fate$result,breaks = seq(-5,5))
fate$dice="fate"
```

```{r pbta dice}
n <- 2
l <- rep(list(1:6), n)

aw=expand.grid(l)
aw$result=rowSums(aw)
aw$dice="aw"


```

```{r, message=FALSE}
ggplot(dice,aes(x=result)) +
  geom_histogram(data=subset(dice,dice == 'fate'),fill = "blue", alpha = 0.2) +
    geom_histogram(data=subset(dice,dice == 'aw'),fill = "red", alpha = 0.2)+
aes(y=stat(count)/sum(stat(count))) +
  scale_y_continuous(labels = scales::percent)#+stat_function(fun = dnorm, args = list(mean = 0, sd = 2))+ stat_function(fun = dnorm, args = list(mean = 0, sd = 3))
```

## Including Plots

You can also embed plots, for example:

```{r dice comparison}
qplot(x =result, data = rbind(fate[,c("result", "dice")], aw[,c("result", "dice")]),geom = "density",color=dice )
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
aw$result=aw$result-7
dice=rbind(fate[,c("result", "dice")], aw[,c("result", "dice")])
# qplot(x =result, data = dice,geom = "density",color=dice, xlim=c(-7,7))

ggplot(dice,aes(x=result)) +
  # geom_density(data=subset(dice,dice == 'fate')) +
    # geom_density(data=subset(dice,dice == 'aw'))+
geom_density(aes(color=dice))+
  stat_function(fun = dnorm, args = list(mean = 0, sd = 2))+ stat_function(fun = dnorm, args = list(mean = 0, sd = 3))+ 
  xlim(c(-7,7))
```




